---
title: Interactions and Nonlinear Relationships
date: today
date-format: long
footer: "[IAFF 6501 Website](https://quant4ia.rocks/)"
logo: images/iaff6501-logo.png
format:
  revealjs:
    theme: [simple, custom.scss]
    transition: fade
    slide-number: true
    #multiplex: true
    chalkboard: true
execute:
  echo: true
  message: false
  warning: false
  freeze: auto
---

```{r}
#| label: setup

library(tidyverse)
library(dplyr)
library(ggplot2)
library(tidymodels)
library(patchwork)
#library(DT)
library(vdemdata)
```

## Oil Wealth and Democracy

Political Resource Curse


## Set up data: Use the year 2005

```{r}
oilData <- vdem %>% 
  filter(year == 2005) %>% 
  select(country_name, v2x_libdem, e_gdppc, v2cacamps, e_total_oil_income_pc, v2x_corr, v2pepwrgen, v2clgencl) %>% 
  mutate(lg_gdppc = log(e_gdppc))
```

## Distribution of Oil Wealth per Capita

```{r echo=FALSE}
oilData %>% 
  ggplot(., aes(x = e_total_oil_income_pc)) +
  geom_histogram(color="lightgrey")
```

## Create a Dummy Variable for High Oil Wealth

```{r}
oilData <- oilData %>% 
  mutate(
        oil = ifelse(e_total_oil_income_pc > 152, 1, 0),
         oil = factor(oil, labels=c("No Oil Wealth", "Oil Wealth"))
        ) %>% 
  filter(oil == "No Oil Wealth" | oil == "Oil Wealth")
table(oilData$oil)

```

## Democracy, GDP per Capita, and Oil Wealth

How should we interpret this graph?

```{r echo=FALSE}
ggplot(oilData, aes(x = lg_gdppc, y = v2x_libdem, color = oil)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "GDP per Capita (log)", y = "Liberal Democracy Index")  +
  scale_color_manual(name=NULL, 
                      values = c("#E99840", "#0C1956"))
```


## Model with Oil and GDP per Capita, No Interaction

What is correct interpretation?

```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(v2x_libdem ~ lg_gdppc + oil, data = oilData) %>% 
  tidy()
```


## Interaction Model

- Does GDP per capita have a different relationship to democracy in Oil rich versus non-Oil rich countries?


## Interaction Model


- $Y_i = a + b_1 * GDPpc + b_2 * Oil + b_3 * GDPpc * Oil$

. . .


- $b_3 * GDPpc * OilWealth$ captures the interaction


## Interaction Model: Interpretation


- How should we interpret $a$?


- $Y_i = a + b_1 * GDPpc + b_2 * Oil + b_3 * GDPpc * Oil$

. . .


- $a$ is predicted level of democracy where GDPpc = 0 and Oil Wealth = 0



## Interaction Model: Interpretation

- What happens if we set Oil Wealth to 0?

- $Y_i = a + b_1 * GDPpc + b_2 * Oil + b_3 * GDPpc * Oil$

- $Y_i = a + b_1 * GDPpc + b_2 * 0 + b_3 * GDPpc * 0$

- $Y_i = a + b_1 * GDPpc$


. . .


- $b_1$ is the association between GDP per capita and democracy, **when oil wealth = 0**


## Interaction Model: Interpretation

- What happens if we set GDPpc to 0?

- $Y_i = a + b_1 * GDPpc + b_2 * Oil + b_3 * GDPpc * Oil$

- $Y_i = a + b_1 * 0 + b_2 * Oil + b_3 * Oil * 0$

- $Y_i = a + b_2 * Oil$

- $b_2$ is the association between oil and democracy, **when GDP per capita = 0**


## Interaction Model: Interpretation

- What happens if we set Oil Wealth to 1?

- $Y_i = a + b_1 * GDPpc + b_2 * 1 + b_3 * GDPpc * 1$

- Rearrange: $Y_i = a + (b_1 + b_3) * GDPpc + b_2 * 1$


. . .

- $b_1 + b_3$ is the association between GDP per capita and democracy, **when oil wealth = 1**


## Run the Interaction Model

What is correct interpretation of each coefficient?

```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(v2x_libdem ~ lg_gdppc*oil, data = oilData) %>% 
  tidy()
```

## Interpretation


$\hat{Y_i}$ = .138 + .164 * $GDPpc_i$ + (-0.085) * $Oil_i$ 
+ (-0.0607) * $GDPpc_i$ * $Oil_i$

. . .

- Relationship without oil: $0.164$

- Relationship with oil: $0.164 + (-0.0607) = 0.164 -0.0607 = 0.1033$
 
## Is this difference due to chance?

- Does GDP have a lower relationship to democracy in oil rich countries?

- Is 0.164 without oil different from 0.1033 with oil?

. . .

- What is the null hypothesis?  What is the alternative?

## Is this difference due to chance?

```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(v2x_libdem ~ lg_gdppc*oil, data = oilData) %>% 
  tidy(conf.int=TRUE) %>% 
  filter(term == "lg_gdppc:oilOil Wealth") %>% 
  select(term, estimate, std.error, p.value, conf.low, conf.high) %>%   
  mutate(across(where(is.numeric), round, 3))
```

## Is this difference due to chance?


- 95% Confidence Interval: [-0.131, 0.009]: mostly negative

- p-value of interaction term: 0.089

    - this is higher that 0.05, but close and remember that the cutoff is arbitrary


## Your Turn: Oil Wealth and Women's Civil Liberties

- V-Dem has a variable capturing respect for women's civil liberties -v2clgencl

- How do GDP per capita and oil wealth relate to women's civil liberties?

- First, use oilData to create a scatterplot that visualizes this relationship.

- Second, fit an interaction model (oil X GDP per capita) predicting women's civil liberties

- Interpret each of the coefficients and the intercept

- What is your overall conclusion(s) from this model?


## Your Turn: Oil Wealth and Women's Civil Liberties

```{r echo=FALSE}
ggplot(oilData, aes(x = lg_gdppc, y =  v2clgencl, color = oil)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "GDP per Capita", y = "Corruption Index")  +
  scale_color_manual(name=NULL, 
                      values = c("#E99840", "#0C1956"))
```

## Your Turn: Oil Wealth and Women's Civil Liberties


```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(v2clgencl ~ lg_gdppc*oil, data = oilData) %>% 
  tidy()
```

