---
title: "Modeling"
subtitle: "Part 3: Multiple Regression"
date: today
date-format: long
footer: "[IAFF 6501 Website](https://data4ia-spring24.netlify.app/)"
logo: images/iaff6501-logo.png
format:
  revealjs:
    theme: [simple, custom.scss]
    transition: fade
    slide-number: true
    #multiplex: true
    chalkboard: true
execute:
  echo: true
  message: false
  warning: false
  freeze: auto
---


## Load packages 

```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(tidymodels)
library(patchwork)
library(DT)
library(vdemdata)
```


## Load VDEM Data

```{r echo=FALSE}
modelData <- vdem %>% 
  filter(year == 2019) %>% 
  select(country_name, v2x_libdem, e_gdppc, v2cacamps, e_total_oil_income_pc, v2x_corr) %>% 
  mutate(lg_gdppc = log(e_gdppc))
```


## Linear Model with Multiple Predictors

- Previously, we were interested in GDP per capita as a predictor of democracy.

- Now, let's consider another predictor: polarization (also measured by V-Dem)


## Polarization Measure in the USA


```{r echo=FALSE}
usa <- vdem %>% 
  filter(country_name == "United States of America") %>% 
  filter(year > 1900) %>% 
  ggplot(., aes(y= v2cacamps, x = year)) +
  geom_line(color = "dodgerblue") +
  labs(y = "Polarization", x = "Year") + ggtitle("Polarization in the USA") 
usa
```



## Polarization and Democracy in 2019

```{r echo=FALSE}
ggplot(modelData, aes(x = v2cacamps, y = v2x_libdem)) +
  geom_point() +
  geom_smooth(method = "lm", color = "#E48957", se = FALSE) +
  labs(x = "Polarization", y = "Liberal Democracy Index") 
```

## Model with Polarization as the only Predictor

How do we interpret the intercept and the slope estimate?

```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(v2x_libdem ~ v2cacamps, data = modelData) %>% 
  tidy()
```


## Model with Two Predictors

<br>

$$\hat{Y_i} = a + b_1*Polarization + b_2*GDPpc$$ 
<br>

- OLS is searching for combination of $a$, $b_1$, and $b_2$ that minimize sum of squared residuals 

    - same logic as model with one predictor, just more complicated


## Model with Two Predictors

```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(v2x_libdem ~ v2cacamps + lg_gdppc, data = modelData) %>% 
  tidy()
```

## Model with Two Predictors

$$\hat{Y_i} = a + b_1*Polarization + b_2*GDPpc$$ 

$$\hat{Y_i} = 0.18 + -0.05*Polarization + 0.10*GDPpc$$ 


## Model with Two Predictors 

$$\hat{Y_i} = a + b_1*Polarization + b_2*GDPpc$$ 

$$\hat{Y_i} = 0.18 + -0.05*Polarization + 0.10*GDPpc$$ 

- $a$ is the predicted level of Y when BOTH GDP per capita and polarization are equal to 0


## Model with Two Predictors 

$$\hat{Y_i} = a + b_1*Polarization + b_2*GDPpc$$ 

$$\hat{Y_i} = 0.18 + -0.05*Polarization + 0.10*GDPpc$$ 

- $b_1$ is the impact of a 1-unit change in polarization on the predicted level of Y, **holding GDP per capita fixed** (all else equal)

    - The relationship between polarization and democracy, controlling for GDP per capita.


## Model with Two Predictors 

$$\hat{Y_i} = a + b_1*Polarization + b_2*GDPpc$$ 

$$\hat{Y_i} = 0.18 + -0.05*Polarization + 0.10*GDPpc$$ 

- $b_2$ is the impact of a 1-unit change in GDP per capita on the predicted level of Y, **holding polarization fixed** (all else equal)


## Your Turn: Polarization, Democracy, and Corruption

- Last week, you examined levels of democracy and corruption

- Today, first fit a multiple regression model predicting corruption with two predictors: democracy (v2x_libdem) and polarization (v2cacamps)

- Interpret the coefficients

- Estimate a second multiple regression model that adds in GDP per capita (lg_gdppc)

- Interpret the coefficients

- What happens to the coefficients on democracy when we add GDP per capita to the model?

    - Why does this happen?



## Model 1

```{r}
mod1 <- linear_reg() %>%
      set_engine("lm") %>%
      fit(v2x_corr ~ v2x_libdem + v2cacamps , data = modelData) %>%
      tidy() %>% 
      mutate(model = "Model 1: No GDP per capita")
mod1
```

## Model 2 adding GDP per capita

```{r}
mod2 <- linear_reg() %>%
      set_engine("lm") %>%
      fit(v2x_corr ~ v2x_libdem + v2cacamps + lg_gdppc, data = modelData) %>%
      tidy() %>% 
      mutate(model = "Model 2: GDP per capita")
mod2
```

<!-- ## Compare -->

<!-- ```{r echo=FALSE} -->
<!-- ms <- rbind(mod1, mod2) -->
<!-- ms %>%  -->
<!--   filter(term !="(Intercept)") %>%  -->
<!--   ggplot(., aes(x = estimate, y = term, color=model)) +  -->
<!--   geom_point(size=2, position=position_dodge(width = -.4)) + -->
<!--     scale_color_manual(name=NULL,  -->
<!--                       values = c("#E99840", "#0C1956")) -->
<!-- ``` -->


# Multiple Predictors: Interaction Models

## Oil Wealth and Democracy

Political Resource Curse


## Set up data: Use the year 2005

```{r}
oilData <- vdem %>% 
  filter(year == 2005) %>% 
  select(country_name, v2x_libdem, e_gdppc, v2cacamps, e_total_oil_income_pc, v2x_corr, v2pepwrgen, v2clgencl) %>% 
  mutate(lg_gdppc = log(e_gdppc))
```

## Distribution of Oil Wealth per Capita

```{r echo=FALSE}
oilData %>% 
  ggplot(., aes(x = e_total_oil_income_pc)) +
  geom_histogram(color="lightgrey")
```

## Create a Dummy Variable for High Oil Wealth

```{r}
oilData <- oilData %>% 
  mutate(
        oil = ifelse(e_total_oil_income_pc > 152, 1, 0),
         oil = factor(oil, labels=c("No Oil Wealth", "Oil Wealth"))
        ) %>% 
  filter(oil == "No Oil Wealth" | oil == "Oil Wealth")
table(oilData$oil)

```

## Democracy, GDP per Capita, and Oil Wealth

How should we interpret this graph?

```{r echo=FALSE}
ggplot(oilData, aes(x = lg_gdppc, y = v2x_libdem, color = oil)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "GDP per Capita (log)", y = "Liberal Democracy Index")  +
  scale_color_manual(name=NULL, 
                      values = c("#E99840", "#0C1956"))
```


## Model with Oil and GDP per Capita, No Interaction

What is correct interpretation?

```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(v2x_libdem ~ lg_gdppc + oil, data = oilData) %>% 
  tidy()
```


## Interaction Model

- Does GDP per capita have a different relationship to democracy in Oil rich versus non-Oil rich countries?


## Interaction Model


- $Y_i = a + b_1 * GDPpc + b_2 * Oil + b_3 * GDPpc * Oil$

. . .


- $b_3 * GDPpc * OilWealth$ captures the interaction


## Interaction Model: Interpretation


- How should we interpret $a$?


- $Y_i = a + b_1 * GDPpc + b_2 * Oil + b_3 * GDPpc * Oil$

. . .


- $a$ is predicted level of democracy where GDPpc = 0 and Oil Wealth = 0



## Interaction Model: Interpretation

- What happens if we set Oil Wealth to 0?

- $Y_i = a + b_1 * GDPpc + b_2 * Oil + b_3 * GDPpc * Oil$

- $Y_i = a + b_1 * GDPpc + b_2 * 0 + b_3 * GDPpc * 0$

- $Y_i = a + b_1 * GDPpc$


. . .


- $b_1$ is the association between GDP per capita and democracy, **when oil wealth = 0**


## Interaction Model: Interpretation

- What happens if we set GDPpc to 0?

- $Y_i = a + b_1 * GDPpc + b_2 * Oil + b_3 * GDPpc * Oil$

- $Y_i = a + b_1 * 0 + b_2 * Oil + b_3 * Oil * 0$

- $Y_i = a + b_2 * Oil$

- $b_2$ is the association between oil and democracy, **when GDP per capita = 0**


## Interaction Model: Interpretation

- What happens if we set Oil Wealth to 1?

- $Y_i = a + b_1 * GDPpc + b_2 * 1 + b_3 * GDPpc * 1$

- Rearrange: $Y_i = a + (b_1 + b_3) * GDPpc + b_2 * 1$


. . .

- $b_1 + b_3$ is the association between GDP per capita and democracy, **when oil wealth = 1**


## Run the Interaction Model

What is correct interpretation of each coefficient?

```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(v2x_libdem ~ lg_gdppc*oil, data = oilData) %>% 
  tidy()
```

## Interpretation


$\hat{Y_i}$ = .138 + .164 * $GDPpc_i$ + (-0.085) * $Oil_i$ 
+ (-0.0607) * $GDPpc_i$ * $Oil_i$

. . .

- Relationship without oil: $0.164$

- Relationship with oil: $0.164 + (-0.0607) = 0.164 -0.0607 = 0.1033$
 
## Is this difference due to chance?

- Does GDP have a lower relationship to democracy in oil rich countries?

- Is 0.164 without oil different from 0.1033 with oil?

. . .

- What is the null hypothesis?  What is the alternative?

## Is this difference due to chance?

```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(v2x_libdem ~ lg_gdppc*oil, data = oilData) %>% 
  tidy(conf.int=TRUE) %>% 
  filter(term == "lg_gdppc:oilOil Wealth") %>% 
  select(term, estimate, std.error, p.value, conf.low, conf.high) %>%   
  mutate(across(where(is.numeric), round, 3))
```

## Is this difference due to chance?


- 95% Confidence Interval: [-0.131, 0.009]: mostly negative

- p-value of interaction term: 0.089

    - this is higher that 0.05, but close and remember that the cutoff is arbitrary


## Your Turn: Oil Wealth and Women's Civil Liberties

- V-Dem has a variable capturing respect for women's civil liberties -v2clgencl

- How do GDP per capita and oil wealth relate to women's civil liberties?

- First, use oilData to create a scatterplot that visualizes this relationship.

- Second, fit an interaction model (oil X GDP per capita) predicting women's civil liberties

- Interpret each of the coefficients and the intercept

- What is your overall conclusion(s) from this model?


## Your Turn: Oil Wealth and Women's Civil Liberties

```{r echo=FALSE}
ggplot(oilData, aes(x = lg_gdppc, y =  v2clgencl, color = oil)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "GDP per Capita", y = "Corruption Index")  +
  scale_color_manual(name=NULL, 
                      values = c("#E99840", "#0C1956"))
```

## Your Turn: Oil Wealth and Women's Civil Liberties


```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(v2clgencl ~ lg_gdppc*oil, data = oilData) %>% 
  tidy()
```