---
title: Authoring in Quarto
subtitle: Advanced HTML
date: today
date-format: long
footer: "[IAFF 6501 Website](https://quant4ia.rocks/)"
logo: images/iaff6501-logo.png
format:
  revealjs:
    theme: [simple, custom.scss]
    transition: fade
    slide-number: true
    #multiplex: true
    chalkboard: true
execute:
  echo: false
  message: false
  warning: false
  freeze: auto
---

## Advanced HTML

<br>

- So far we have been using the basic HTML elements to create our documents
- Quarto supports a wide range of HTML elements and attributes
- We can use these to create more complex and interactive documents

## Formatting Options

- We can make our documents have a more sophisticated feel by specifying various options in the YAML header
- Examples include `title`, `subtitle`, `date`, `abstract`, `toc`, etc. 
- We can also specify font styles and colors and control other aspects of the formatting
- See [this page](https://quarto.org/docs/reference/formats/html.html) to explore the various formatting options

## Themes

<br>

- Quarto supports a wide range of themes
- We can use these to change the look and feel of our documents
- Check out [this page](https://quarto.org/docs/output-formats/html-themes.html) to explore the various themes available in Quarto

## Generating Lorem Ipsum

<br>

- For working on your document's format it is helpful to have random text
- The [lorem package](http://pkg.garrickadenbuie.com/lorem/) is helpful for this
- You can also install an R Studio Addin

## Your Turn! {.smaller}

- Start a new Quarto project
- Create a new Quarto document in your project folder
- Generate some sections and some random text using the `lorem` package
- Add a YAML header with the following options:
  - `title:`
  - `subtitle:`
  - `date:`
  - `date-format:`
  - `theme:`
  - `toc:`
- Try adding at least one more option to your YAML header 
- Render the document and see what it looks like

```{r}
#| label: timer1
library(countdown)
countdown(minutes = 10, 
          id = "timer1", 
          bottom = "10%", 
          right = "10%",
          color_border = "#fff",
          color_text = "#fff",
          color_running_background = "#42affa",
          color_running_text = "black",
          color_finished_background = "#E5D19D",
          color_finished_text = "#00264A")
```

## Citations

<br>

- With Quarto you can easily add citations to your document
- Then a bibliography will be automatically generated
- Too add a citation you can use the `@` symbol followed by the citation key
- In order to do this you need to know a little bit about BibTeX and/or CSL

## BibTeX

<br>

- BibTeX is a reference management software package
- When we retrieve a citation for use in Quarto it should be in BibTeX format
- BibTeX is a plain text format that is easy to read and write
- To work with BibTeX you need a reference manager like Zotero or Mendeley
- We are going to do Zotero

## Zotero

<br>

- Go to [Zotero](https://www.zotero.org/)
- Create an account
- Download the Zotero app
- Install the Zotero Connector for your browser
- Install the Better BibTeX extension
  - Follow [these instructions](https://retorque.re/zotero-better-bibtex/installation/)
  
## Downloading Citations

<br>

- Open Zotero
- Go to google scholar
- Search for a paper on a topic related to your research
- Click the Zotero Connector to download the citation

## Adding Citations

<br>

- Export the bibliography as a .bib file from Zotero
- Add the .bib file to your Quarto project folder
- Add `bibliography:` and the name of the file to your YAML header
  - e.g. `bibliography: mybib.bib`

## Citations in Text

<br>

- Now you can add citations to your text
- Use the `@` symbol followed by the citation key
- For example `@smith2020` will add a citation to the paper by Smith (2020) assuming you have a citation in your .bib file with the key `smith2020`
- Now you can render your document and see the bibliography at the end

## Your Turn! 

<br>

- Add a bibliography to your Quarto document
- Add a citation to your text
- Render the document and see the bibliography at the end

```{r}
#| label: timer2
library(countdown)
countdown(minutes = 10, 
          id = "timer2", 
          bottom = "10%", 
          right = "10%")
```

## Footnotes

<br>

- You can also add footnotes to your document
- Use the `^[]` or `[^1]` syntax to add a footnote
- There are different ways to add footnotes
- I find that inline notes are the easiest to work with
  - For example `^[This is a footnote]`
  
  
## Scholarly Writing

<br>

- Quarto is a great tool for scholarly writing
- There are lots of formatting options for scholarly papers specifically
- Check out [this section](https://quarto.org/docs/authoring/front-matter.html) of the Quarto guide for more details

## Publishing on Quarto Pub

<br>

- [Quarto Pub](https://quartopub.com/) is a platform for publishing Quarto documents
- You can publish your documents on Quarto Pub and share them with others
- Quarto Pub is free to use
- Sign up for an account, then publish with `quarto publish quarto-pub name-of-document.qmd` in the terminal

## Your Turn!

<br>

- Add a footnote to your Quarto document
- Explore the scholarly writing options in Quarto
- Publish your document on Quarto Pub

```{r}
#| label: timer3
library(countdown)
countdown(minutes = 10, 
          id = "timer3", 
          bottom = "10%", 
          right = "10%")
```


